---
title: "HW2 - EDA with R - Part 1: Statistics and Plotting"
author: "misken"
date: "January 9, 2022"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Step 1 - Familiarize yourself with the data and the assignment

In this assignment you'll do some exploratory data analysis
with R on a dataset of housing prices. The version of
this data that we'll be using is based on a Kaggle dataset -
https://www.kaggle.com/harlfoxem/housesalesprediction. I've made 
some changes and you can find a data dictionary in the same
folder as this file.

As we did in the R related sessions, you'll be editing an R Markdown document to
do some analysis as well as to document the
steps you did (and answer some questions I'll throw at you).

You'll notice a few "Hacker Extra Credit" tasks
thrown in. These are for those of you who want to go a little above and beyond
and attempt some more challenging tasks. And, feel free to do a some free form
Hacker Extra style work yourself - in other words, do something beyond what
was asked. You'll learn more.

## Step 2 - Create a new R Markdown document based on this document

Save this file as a new R Markdown document and name it **hw2_eda_part1_[your last name].Rmd**. Mine would
be called **hw2_eda_part1_isken.Rmd**. Save it into the same folder as this file. 

This will be your working directory. 

Create a new R Project based on this working directory.

## Step 3 - Explore data folder

Inside of the main project folder you'll find the data files for this assignment in a subfolder called `data/`.

- **kc_house_data_original.csv**
- **kc_house_data.Rdata**


The first file is from the Kaggle site for this dataset. I made some changes and created the Rdata file which contains the dataframe that you'll be using for your analysis.


You might want to use shell commands or a text editor like Geany to snoop the csv file a bit.


## Step 4 - Complete the following R tasks and answer questions

Now you'll need to complete the following tasks in R. Just like we did in class, you should
use a combination of markdown text (be concise, no need to write tons of text) 
to explain what you are doing and R code chunks to actually do it.
When you are done, use the "Knit" button to generate an HTML file from your R Markdown.

Note that if you have errors in your R code, **knitr** will stop at the error
and won't create an html file. Sooo, you should comment out R code that has
errors and write a little comment stating that it's an error and a possible explanation for the error. Of course, try to fix the errors. Keep doing this
until you successfully knit an html file based on your R Markdown document.
You'll be submitting both the html file as well as the Rmd file.

You'll notice that I've left some empty code chunks
as well as some partially filled in code chunks. Of course, feel free to
add more code chunks as needed.

Let's load a few libraries we'll need:

```{r}
library(dplyr)
library(ggplot2)
```


### Problem 1 - Load the Rdata file

Let's load the data from the Rdata file.

```{r load_data}
load("data/kc_house_data.Rdata")

```


Check the structure of the data.frame using `str` and use the `summary` function to begin exploring the data. 

```{r str_summary}
str(kc_house_data)
summary(kc_house_data)

```

**QUESTION** What does each row represent?

> Each row provides a vector on the properties ()unique identifier, date of sale, price, number of bedrooms etc.) of each home sold. 

**QUESTION** Are there any NA values? 

```{r}
sum(is.na(kc_house_data))
```


> There are no NA values



**Unless otherwise specified, you should use the ggplot2 package for all of the following plots.**

### Problem 2 - Basic plots

Create a histogram for `price` and for `sqft_living`.

```{r}
# Histogram for price
ggplot(kc_house_data) +
  geom_histogram(aes(x=price))


#Histogram for sqft_living
ggplot(kc_house_data) +
  geom_histogram(aes(x=sqft_living))

```


In addition, compute and display the following statistics for price and for sqft_living:

* mean and median
* standard deviation
* min, max and range
* 95th and 99th percentile
* IQR
* coefficient of variation

There are few different ways to print out a string that contains some literal text along with the value of some variable. For example, let's compute the mean of `price` and store it in a variable.

```{r}
# mean
price_mean <- mean(kc_house_data$price)
price_mean
sqft_living_mean <- mean(kc_house_data$sqft_living)
sqft_living_mean

# median
price_median <- median(kc_house_data$price)
price_median
sqft_living_median <- median(kc_house_data$sqft_living)
sqft_living_median

# Standard Deviation
price_stddev <- sd(kc_house_data$price)
price_stddev
sqft_living_stddev <- sd(kc_house_data$sqft_living)
sqft_living_stddev

# min
price_min <- min(kc_house_data$price)
price_min
sqft_living_min <- min(kc_house_data$sqft_living)
sqft_living_min

# max
price_max <- max(kc_house_data$price)
price_max
sqft_living_max <- max(kc_house_data$sqft_living)
sqft_living_max

# range
price_range <- range(kc_house_data$price)
price_range
sqft_living_range <- range(kc_house_data$sqft_living)
sqft_living_range

# 95th percentile
price_95percentile <- quantile(kc_house_data$price, 0.95)
price_95percentile
sqft_living_95percentile <- quantile(kc_house_data$sqft_living, 0.95)
sqft_living_95percentile

# 99th percentile
price_99percentile <- quantile(kc_house_data$price, 0.99)
price_99percentile
sqft_living_99percentile <- quantile(kc_house_data$sqft_living, 0.99)
sqft_living_99percentile

# IQR
price_IQR <- IQR(kc_house_data$price)
price_IQR
sqft_living_IQR <- IQR(kc_house_data$sqft_living)
sqft_living_IQR

#coefficient of variation
price_cv <- price_stddev/price_mean
price_cv
sqft_living_cv <- sqft_living_stddev/sqft_living_mean
sqft_living_cv

```

If I wanted to create a little message such as "Mean price = 54088.1", I could use a few different approaches. An old standby is the R `paste()` function.

```{r}
paste("Mean price =", price_mean)

```

A few things to note:

* by default, `paste` will concatenate the list of arguments AND put a space between them (see help(paste) to control the delimiter)
* the value prints in a default format

Another approach is to use the venerable `sprintf()` function, which is actually based on "C-style printing" from the C language. A nice intro to this is at https://www.gastonsanchez.com/r4strings/c-style-formatting.html.

```{r}
sprintf("Mean price = %.2f", price_mean)
```

If you wrap this with `cat()` function, you can get rid of the output number prefix.

```{r}
cat(sprintf("Mean price = %.2f", price_mean))
```

```{r summary_stats}

```


```{r histos}

```


Write a short paragraph describing the distribution of of these two variables.

> PUT ANSWER HERE

Below I demo how you can include *inline code* to make even your summary comments
more dynamic. These inline code snippets get run during knitting.

Repeat the histogram for `price` but make the following style changes to the plot.

* Add an approprate graph title 
* Change y-axis label to "Number of houses"
* Change x-axis label to "Selling price"
* Change the colors of the bars to something other than the default

```{r histo_stylechanges}

```

Now create a set of histograms for `price` in which you facet by condition

```{r histos_price}
# ggplot(???) + 
#   geom_histogram(aes(x=???)) +
#   ???
```


Since the number of houses varies by `condition`, create a new version of the plot above but modify so that the y-axis shows frequency percentages (or density values) instead of
counts. Hint: See the [documentation](https://ggplot2.tidyverse.org/reference/geom_histogram.html)

```{r histos_condition_density}

```

Now create boxplots and violin plots for the same two variables - `price` and `sqft_living`. However, you'll
see that for both of these plot types, ggplot wants an `x` aesthetic to act as a grouping variable. Use the `condition` field as your `x` aesthetic. Note anything of interest revealed by these plots.

```{r box_violin}
# Boxplots and violin plots

```


> PUT SUMMARY COMMENTS HERE


Repeat the histogram of `sqft_living` but only for houses with a `condition` equal to 3. You can either use
dplyr to do the filtering or boolean indexing (see intro to dataframes Rmd file). 

**HACKER EXTRA:** In addition, overlay a kernel density plot on top of the histogram.

```{r histo_cond3}



```



Create boxplots and/or violin plots showing how prices compare with respect to the number of bedrooms.

Summarize your findings.

```{r boxplots_bedrooms}


```

> PUT YOUR SUMMARY COMMENTS HERE

### Problem 3 - Basic bar chart

Create a bar chart showing the number of houses by grade. 

```{r bar_grade}


```


### Problem 4 - Basic scatter plots

Create a scatter plot of prices vs sqft_living.

```{r xy_price_sqft}

```


Create a modified version of the above scatter plot in which:

* only houses less than 3000 square feet are included,
* the point color is mapped to condition 

You could also facet by condition if you think that makes the plot easier to interpret.

```{r xy_price_sqft_condition}

```

> PUT YOUR SUMMARY COMMENTS HERE

Make one more scatter plot involving `price` and `sqft_living`. But, figure out a way to either change the axis scales or 
the plotted data itself to use a base 10 logarithmic scale (or transform).
Write a short paragraph on why one might want to use a log scale or log
transform of data when plotting.



### Problem 5 - correlations

Create a correlation matrix and a correlation plot (using corrplot library) which includes the following variables:

* price (col 3)
* bedrooms (col 4)
* bathrooms (col 5)
* sqft_living (col 6)
* sqft_lot (col 7)
* floors (col 8)
* sqft_above (col 13)
* sqft_basement (col 14)
* sqft_living15 (col 20)
* sqft_lot15 (col 21)
* age (col 22)

```{r correlation_matrix}

```

```{r correlation_plot}

```


> Discuss your findings.



### Problem 6 - automated EDA

There are a number of packages that try to help automate the process of initial 
EDA of a dataframe. I want you to try out the skimr package. If the package
isn't already installed, just install it. Run the `skim()` function on `kc_house_data` and then write a few short paragraphs describing the types of outputs you get from skimr.

```{r lib_skimr}
library(skimr)
```

```{r skim}
# Use skim() function on kc_house_data

```


## Deliverables for this problem

Make sure you knit your Rmd file. IMPORTANT: You might notice that there's a file with the same base name as your file and with a .nb.html extension. The "nb" stands for "notebook" and that file is created on the fly as you are working in your notebook. DO NOT SUBMIT THAT FILE. YOU MUST DO A "knit to html" from the toolbar which will generate a file with just the .html extension. As discussed earlier, if you have errors, your document won't knit. Comment out error lines and put a comment in saying that you couldn't get the code chunk to work. After your file is knitted into an html file, simply create a compressed archive (either tar.gz or .zip) of your project folder and upload the compressed archive file into Moodle.